Planning to Save Energy: How Information Format Affects Accuracy

### 2025-03-17 manuscript/intro.qmd
writing updates from last week I forgot to push
Update after fixing some apa issues; grammatical errors etc.

@@

>>>>

---
title: "Planning to Save Energy: How Information Format Affects Accuracy"
author:
  - name: Thomas E. Gorman
    url: https://tegorman13.github.io/
    orcid: "0000-0001-5366-5442"
    affiliations: 
      - ref: cclab
    attributes: 
      corresponding: true
affiliations:
  - id: cclab
    name: Communication and Cognition Lab, Purdue University
    city: West Lafayette
    region: IN
    country: USA
    postal-code: 47907
    url: https://web.ics.purdue.edu/~treimer/
date: 2025-03-16
abstract: "Effective communication of energy consumption information is crucial for promoting residential energy conservation. This study investigates how different numerical representations of energy reduction goals influence consumers' ability to create accurate conservation plans. Across two experiments, we examined the impact of presenting energy information in kilowatt-hours (kWh), percentages, or U.S. dollars (USD) on planning accuracy. Participants completed a simulated household planning task in which they allocated energy usage across multiple appliances, with the goal presented in either kilowatt-hours (kWh), percentages, or monetary costs. Results across both experiments showed that presenting reduction goals in absolute units (kWh) led to significantly greater accuracy compared to percentage-based or monetary formats. Furthermore, we found that higher energy literacy was associated with more accurate planning. These findings demonstrate that absolute units (kWh) are more effective for communicating energy-saving goals, and highlight the potential value of educational interventions to improve consumer energy literacy."
bibliography: ../assets/rs_bib.bib
csl: ../assets/apa.csl
keywords: ["Information format", "Energy literacy", "Decision-making", "Planning"]
floatsintext: true
---

@@

::: {#tbl-s1-ord apa-note="Odds ratios greater than 1 indicate increased odds of falling into a worse accuracy category compared to the kWh condition"}
|Comparison        |odds_ratio |ci_lower |ci_upper |
|:-----------------|:----------|:--------|:--------|
|Percentage vs kWh |4.2        |1.1      |18       |
|USD vs kWh        |22.9       |6.1      |90       |

**Experiment 1**: Odds ratios for group comparisons

:::

For1 each family scenario, the participants were shown a table containing the families utility usage from the prior year, alongside the state averages for each appliance category (both prior year usage and stage averages are always shown in kWh). For each scenario, participants were asked to create two possible action plans to achieve the target reduction in total household energy usage (see fig-task). Depending on their reference class condition, the target reduction amount presented either in kilowatt-hours (kWh), as percentages of total household usage, or in U.S. dollars. In all conditions, the target reduction was equivalent to a 15% reduction in total household kWh.

@gigerenzerSimpleToolsUnderstanding2003

![Example energy planning task trial](/assets/images/wells10e_pub.png){#fig-task apa-note="Participants saw a table with a family's previous year electricity usage (here for the Wells family in Colorado) and were asked to allocate energy usage to meet a 15% reduction goal. The format of the reduction goal was manipulated to be either a percentage (15% given as goal reduction), kilowatt hours (5965 kWh given), or USD (\$656). Participants in the USD condition were provided with the conversion rate between kwH and USD."}

For2 each family scenario, the participants were shown a table containing the families utility usage from the prior year, alongside the state averages for each appliance category (both prior year usage and stage averages are always shown in kWh). For each scenario, participants were asked to create two possible action plans to achieve the target reduction in total household energy usage (see fig-task). Depending on their reference class condition, the target reduction amount presented either in kilowatt-hours (kWh), as percentages of total household usage, or in U.S. dollars. In all conditions, the target reduction was equivalent to a 15% reduction in total household kWh.

::: {#tbl-s1-agg apa-note="The table shows performance as both the % of trials where participants matched the goal, and the mean absolute error from the target reduction goal. Percentage meeting goal refers to … ETC"}

|Reference Class |Avg. % Change |% meeting goal (exact) |% meeting goal (close match) |Abs. Deviation |Log Abs. Deviation |
|:---------------|:-------------|:----------------------|:----------------------------|:--------------|:------------------|
|kWh             |0.22          |0.38                   |0.54                         |0.03           |-3.7               |
|Percentage      |0.21          |0.22                   |0.40                         |0.06           |-3.1               |
|USD             |0.23          |0.10                   |0.22                         |0.10           |-2.4               |

Study 1: Summary of planning accuracy by reference class

:::

@tbl-s1-agg shows the average x,y,z, separately for the conditions of … . As can be seen in Table 1, participants in the kWh condition met the target goal 38% of the time, compared to 22% for the Percentage condition and 10% for the USD condition. Moreover, the kWh reference class exhibited smaller deviations from the target reduction, suggesting that participants performed more accurately when the goal was framed in kWh rather than when the goal was framed in percentages or USD.

As shown in @tbl-s1-agg, participants in the kWh condition exactly met the target reduction goal 38% of the time, significantly outperforming those in the Percentage (22%) and USD (10%) conditions. Furthermore, the kWh reference class exhibited notably smaller mean absolute deviations (0.03) compared to Percentage (0.06) and USD (0.10), suggesting that presenting the reduction goal in absolute units facilitated more precise allocations.

### 2025-03-17 manuscript/manuscript.qmd
Update after fixing some apa issues; grammatical errors etc.

---
title: "Planning to Save Energy: How Information Format Affects Accuracy"
aliases:
  - /manuscript.html
format-links:   
    - text: HTML
      href: manuscript.html
      icon: file-minus
    - text: HTML (with code)
      href: manuscript-code.html
      icon: file-plus
    - format: apaquarto-pdf
      text: PDF version
      icon: file-pdf
    - text: docx
      href: rs_manuscript_apa.docx
    - gfm
date: 2025-01-25
abstract: "Effective communication of energy consumption information is crucial for promoting residential energy conservation. This study investigates how different numerical representations of energy reduction goals influence consumers' ability to create accurate conservation plans. Across two experiments, we examined the impact of presenting energy information in kilowatt-hours (kWh), percentages, or U.S. dollars (USD) on planning accuracy. Participants completed a simulated household planning task in which they allocated energy usage across multiple appliances, with the goal presented in either kilowatt-hours (kWh), percentages, or monetary costs. Results across both experiments showed that presenting reduction goals in absolute units (kWh) led to significantly greater accuracy compared to percentage-based or monetary formats. Furthermore, we found that higher energy literacy was associated with more accurate planning. These findings demonstrate that absolute units (kWh) are more effective for communicating energy-saving goals, and highlight the potential value of educational interventions to improve consumer energy literacy."
---

@@

Given the relevance of promoting behavior change to reduce energy consumption, several strands of research have aimed to identify factors that affect energy behaviors and interventions that effectively reduce energy consumption. Interventions often focus on X, Y, Z. For example, …

@@

Receiving feedback about one’s energy use is an important step in identifying potential actions to reduce energy consumption. However, it is crucial to understand how consumers use the information that is given with their energy bill and whether consumers are able to translate a energy saving goal into an action plan.

@@

The way numerical information is presented can significantly affect how individuals process and use that information. Of particular relevance are reference class effects, which occur when numerical statements are presented without a clear or intuitive basis for comparison, making it difficult to infer meaningful quantities[@gigerenzerSimpleToolsUnderstanding2003; @reimerNumericCommunicationRisk2015]. A substantial body of evidence suggests that presenting data in terms of absolute counts or frequencies, as opposed to probabilities or percentages, can promote more accurate comprehension and facilitate decision making. However, it is important to note that even intuitive formats can pose challenges. For instance, @weberWhyCanOnly2018 found that individuals often struggle with reasoning tasks presented in natural frequencies because they inadvertently revert to more complex probabilistic thinking.

@@

Despite existing studies on energy-use communication and format effects, limited research has explored how different numerical representations influence consumers' ability to create accurate energy conservation plans. Specifically, there is a gap in understanding how presenting energy information in absolute units versus percentages or monetary terms affects the precision of planning appliance-specific reductions. The current study addresses these critical issues by systematically investigating the impact of varying information formats (kWh, percentage, and USD) on the accuracy of energy-planning decisions. By manipulating the presentation format of energy information, this research aims to elucidate how different representational formats influence planning accuracy. Based on the literature reviewed, we hypothesize that: 1) Presenting energy reduction goals in absolute units (kWh) will lead to greater planning accuracy compared to percentage-based or monetary formats, as absolute units provide a more direct and less ambiguous representation of energy quantities. 2) Higher energy literacy will be associated with more accurate planning, as individuals with greater energy knowledge may be better equipped to process and utilize the provided information, regardless of format. We also examine the potential of several exploratory variables, such as goal difficulty and the rounding of numerical values, to further elucidate the factors that influence planning accuracy.

@@

```{r}
#| label: fig-s1-log-dist
#| fig-cap: "Experiment 1: Planning Error Distributions"
#| apa-note: "Distribution of the log of the absolute error between the participant's action plan and the reduction goal across different reference class conditions (kWh, Percentage, USD). The right side plots are further separated by calculator usage. A lower log absolute error suggests higher planning accuracy."
#| fig-width: 11
#| fig-height: 8

@@

```{r}
#| label: tbl-s1-prop
#| tbl-cap: "Study 1: Descriptive Performance"
#| apa-note: "The table shows the percentage of participants who fell into each accuracy level for each reference class condition (percentages of kWh, $, and USD columns reflect within condition percentages). The combined group column reflects the percentage of participants in each accuracy level when aggregating across across all reference class conditions."

@@

This approach allowed us to estimate thresholds (intercepts) and regression coefficients that capture how different reference classes affect the likelihood of achieving higher accuracy categories. For each comparison, we provide posterior odds ratios (OR) and their 95% CIs. This approach allows the estimation of threshold parameters and regression coefficients that characterize how changes in predictor variables (such as the reference class: kWh, percentage, or USD) relate to probabilities of being in each accuracy category. Specifically, we used a cumulative logit link function to model the ordered accuracy outcome, and we specified weakly informative priors for the regression coefficients (normal distributions with mean 0 and standard deviation of 1) and for the cutpoints (normal distributions with a mean of zero and a standard deviation of 4.0). The approach allows us to estimate threshold parameters and regression coefficients that characterize how changes in predictor variables (such as the reference class: kWh, percentage, or USD) relate to probabilities of being in each accuracy category.

@@

**Experiment 1**: Ordinal Regression results

@@

::: {#tbl-s2-agg apa-note="The table shows performance as both the % of trials where participants matched the goal, and the mean absolute error from the target reduction goal."}

@@

```{r}
#| label: fig-s2-log-dist
#| fig-cap: "Experiment 2: Planning Error Distributions"
#| apa-note: "Distribution of the log of the absolute error between the participant's action plan and the reduction goal across different reference class conditions (kWh, Percentage, USD). The right side plots are further separated by calculator usage. A lower log absolute error suggests higher planning accuracy."
#| fig-width: 11
#| fig-height: 8

>>>>

---
title: "Planning to Save Energy: How Information Format Affects Accuracy"
aliases:
  - /manuscript.html
format-links:   
    - text: HTML
      href: manuscript.html
      icon: file-minus
    - text: HTML (with code)
      href: manuscript-code.html
      icon: file-plus
    - format: apaquarto-pdf
      text: PDF version
      icon: file-pdf
    - text: docx
      href: rs_manuscript_apa.docx
    - gfm
date: 2025-03-16
abstract: "Effective communication of energy consumption information is crucial for promoting residential energy conservation. This study investigates how different numerical representations of energy reduction goals influence consumers' ability to create accurate conservation plans. Across two experiments, we examined the impact of presenting energy information in kilowatt-hours (kWh), percentages, or U.S. dollars (USD) on planning accuracy. Participants completed a simulated household planning task in which they allocated energy usage across multiple appliances, with the goal presented in either kilowatt-hours (kWh), percentages, or monetary costs. Results across both experiments showed that presenting reduction goals in absolute units (kWh) led to significantly greater accuracy compared to percentage-based or monetary formats. Furthermore, we found that higher energy literacy was associated with more accurate planning. These findings demonstrate that absolute units (kWh) are more effective for communicating energy-saving goals, and highlight the potential value of educational interventions to improve consumer energy literacy."
---

@@

Given the relevance of promoting behavior change to reduce energy consumption, several strands of research have aimed to identify factors that affect energy behaviors and interventions that effectively reduce energy consumption. For example, @abrahamseEffectTailoredInformation2007 used a combination of tailored information about energy use and tailored feedback, in addition to setting an energy-saving goal, to promote direct and indirect energy-saving behaviors and high levels of knowledge regarding energy use. Direct energy-saving behaviors were classified as reducing fuel, gas, and electricity consumption, while indirect behaviors referred to producing, distributing, and disposing goods. After a five-month intervention, the authors found that households in the experimental group reduced their direct energy use by 5.1% and had higher levels of knowledge about energy conservation compared to the control group, which used 0.7% more energy since the beginning of the intervention. No difference between groups was found in indirect behaviors.

@@

Receiving feedback about one’s energy use is important in identifying potential actions to reduce energy consumption. However, it is crucial to understand how consumers use the information provided with their energy bill and whether they are able to translate an energy-saving goal into an action plan. @canfieldPerceptionsElectricityuseCommunications2017 specifically studied this problem by running an experiment in which participants were shown hypothetical electricity bills with information related to a household’s historical electricity use, electricity use in relation to their neighbors, and historical electricity use by appliances. Participants were randomly assigned to one of three formats of information representation (i.e., tables, bar graphs, and icon graphs) and were asked questions regarding their energy literacy. @canfieldPerceptionsElectricityuseCommunications2017 showed that tables were the easiest format to understand for consumers when evaluating every type of information related to energy use. Across types of information, historical electricity use elicited the highest intentions and preferences for energy savings regardless of format. Additionally, participants with high energy literacy had a better understanding of energy-related information across all types of information representations. By disentangling the effects of content, format, and individual differences in energy literacy on understanding, preferences, and intentions, @canfieldPerceptionsElectricityuseCommunications2017 demonstrated that easy-to-implement communication strategies in energy bills can lead to energy-saving behaviors.

@@

The way numerical information is presented can significantly affect how individuals process and use that information. Of particular relevance are reference class effects, which occur when numerical statements are presented without a clear or intuitive basis for comparison, making it difficult to infer meaningful quantities [@gigerenzerSimpleToolsUnderstanding2003; @reimerNumericCommunicationRisk2015]. A substantial body of evidence suggests that presenting data in terms of absolute counts or frequencies, as opposed to probabilities or percentages, can promote more accurate comprehension and facilitate decision making. However, it is important to note that even intuitive formats can pose challenges. For instance, @weberWhyCanOnly2018 found that individuals often struggle with reasoning tasks presented in natural frequencies because they inadvertently revert to more complex probabilistic thinking.

@@

Despite existing studies on energy-use communication and format effects, limited research has explored how different numerical representations influence consumers’ ability to create accurate energy conservation plans. Specifically, there is a gap in understanding how presenting energy information in absolute units versus percentages or monetary terms affects the precision of planning appliance-specific reductions. The current study addresses these critical issues by systematically investigating the impact of varying information formats (kWh, percentage, and USD) on the accuracy of energy-planning decisions. By manipulating the presentation format of energy information, this research aims to elucidate how different representational formats influence planning accuracy. Based on the literature reviewed, we hypothesize that: 1) Presenting energy reduction goals in absolute units (kWh) will lead to greater planning accuracy compared to percentage-based or monetary formats, as absolute units provide a more direct and less ambiguous representation of energy quantities. 2) Higher energy literacy will be associated with more accurate planning, as individuals with greater energy knowledge may be better equipped to process and utilize the provided information, regardless of format. We also examine the potential of several exploratory variables, such as goal difficulty and the rounding of numerical values, to further elucidate the factors that influence planning accuracy.

@@

```{r}
#| label: fig-s1-log-dist
#| fig-cap: "Experiment 1: Planning Error Distributions"
#| apa-note: "Distribution of the log of the absolute error between the participant's action plan and the reduction goal. Left side plot separates across different reference class conditions (kWh, Percentage, USD). Right side plot separates by self-reported calculator use. A lower log absolute error suggests higher planning accuracy."
#| fig-width: 11
#| fig-height: 8

@@

```{r}
#| label: tbl-s1-prop
#| tbl-cap: "Experiment 1 1: Categorization of Participants according to Accuracy Levels"
#| apa-note: "The table shows the percentage of participants who fell into each accuracy level for each reference class condition (percentages of kWh, $, and USD columns reflect within condition percentages). The combined group column reflects the percentage of participants in each accuracy level when aggregating across across all reference class conditions."

@@

This approach allowed us to estimate thresholds (intercepts) and regression coefficients that capture how different reference classes affect the likelihood of achieving higher accuracy categories. For each comparison, we provide posterior odds ratios (OR) and their 95% CIs. This approach allows the estimation of threshold parameters and regression coefficients that characterize how changes in predictor variables (such as the reference class: kWh, percentage, or USD) relate to probabilities of being in each accuracy category. Specifically, we used a cumulative logit link function to model the ordered accuracy outcome, and we specified weakly informative priors for the regression coefficients (normal distributions with mean 0 and standard deviation of 1) and for the cutpoints (normal distributions with a mean of zero and a standard deviation of 4.0). The approach can be used to estimate threshold parameters and regression coefficients that characterize how changes in predictor variables (such as the reference class: kWh, percentage, or USD) relate to probabilities of being in each accuracy category.

@@

**Experiment 1**: Ordinal Regression Results of a Test of Differences between conditions in Accuracy Levels

@@

::: {#tbl-s2-agg apa-note="The table shows performance as both the % of trials where participants matched the goal, and the mean absolute error from the target reduction goal. Percentage meeting goal refers to participants who exactly met the target reduction goal (i.e., the proposed plan would reduce energy use by exactly 15%)"}

@@

```{r}
#| label: fig-s2-log-dist
#| fig-cap: "Experiment 2: Planning Error Distributions"
#| apa-note: "Distribution of the log of the absolute error between the participant's action plan and the reduction goal. Left side plot separates across different reference class conditions (kWh, Percentage, USD). Right side plot separates by self-reported calculator use. A lower log absolute error suggests higher planning accuracy."
#| fig-width: 11
#| fig-height: 8
